# Chapter 6 - Analysis of longitudial data  
# Part 1


## Reading the files and checking their contents

```{r}
# Load required packages.
library(dplyr)
library(lme4)
library(ggplot2)
library(tidyr)

# Reading data in.
bprs_long <- read.csv("data/bprs_long.csv")
rats_long <- read.csv("data/rats_long.csv")

# Just for measure, changing the variables to factor format and sorting by week / Time.
bprs_long$treatment <- factor(bprs_long$treatment)
bprs_long$subject <- factor(bprs_long$subject)

bprs_long <- arrange(bprs_long, week)

rats_long$Group <- factor(rats_long$Group)
rats_long$ID <- factor(rats_long$ID)

rats_long <- arrange(rats_long, Time)
```
  
## Working with the RATS dataset

```{r}
# Check colnames, dimensions and structure
colnames(rats_long)
dim(rats_long)
str(rats_long)
```
  The rats dataset (long format) contains The variables ID, Group, Time and weight.  
It has 176 rows and 4 columns.  
ID and Group variables are factor types, Time and weight integers.

## Plotting the data

```{r}
# Drawing the plot.

ggplot(rats_long, aes(x = Time, y = weight, group = ID)) +
  geom_line(aes(linetype = Group, colour = Group)) +
  scale_y_continuous(limits = c(min(rats_long$weight), max(rats_long$weight)))

glimpse(rats_long)
# Standardise the variable weight.
rats_std <- rats_long %>%
  group_by(Time) %>%
    mutate(stdweight = (weight - mean(weight))/sd(weight)) %>%
  ungroup()

# Glimpse the data
glimpse(rats_std)

# Plot again with the standardised bprs

ggplot(rats_std, aes(x = Time, y = stdweight, group = ID)) +
  geom_line(aes(linetype = Group, colour = Group)) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(rats_std$stdweight), max(rats_std$stdweight)))
```
  We can see that the rats in groups 2 and 3 gain more weight as time passes, and have more weight in the beginning than group 1 rats. However, and even more clearly after the standardization, we can see that the group 1 rats do not gain as much weight as the other two.
  
## Summary graphs
```{r}
# Work with the exercise in this chunk, step-by-step. Fix the R code!
# BPRSL is available

# Number of subjects (per group):

sum(rats_std$Group == 1)
sum(rats_std$Group == 2)
sum(rats_std$Group == 3)

n = 44

# Summary data with mean and standard error of bprs by treatment and week
rats_std <- rats_std %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(weight), se =  sd(weight)/sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(rats_std)

# Plot the mean profiles
library(ggplot2)
ggplot(rats_std, aes(x = Time, y = mean, col = Group)) +
  geom_line() +
  #geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")

```
  From the mean profile plots, we can see that the rats do gain weight over time. The rate is faster for groups 2 and 3 as previously observed. There is also an interesting dip in weights for the groups 2 and 3 at around 37-42 days which is not true for group 1.  

## Finding the outliers

```{r}
# Creating a summary data with grouping Group and ID with weight mean as summary variable.
rats_summary <- rats_long %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(weight)) %>%
  ungroup()

# Checking the data with glimpse.
glimpse(rats_summary)

# Drawing the boxplot of mean values between groups.
ggplot(rats_summary, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), Time (days) 1-64")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data

# Splitting the data by rat groups, removing the outliers for each group and combining them with rbind.
rats1 <- rats_summary[rats_summary$Group == 1,]
rats2 <- rats_summary[rats_summary$Group == 2,]
rats3 <- rats_summary[rats_summary$Group == 3,]

g1_Q1 <- quantile(rats1$mean, .25)
g1_Q3 <- quantile(rats1$mean, .75)
g1_IQR <- IQR(rats1$mean)

g2_Q1 <- quantile(rats2$mean, .25)
g2_Q3 <- quantile(rats2$mean, .75)
g2_IQR <- IQR(rats2$mean)

g3_Q1 <- quantile(rats3$mean, .25)
g3_Q3 <- quantile(rats3$mean, .75)
g3_IQR <- IQR(rats3$mean)

rats1 <- subset(rats1, rats1$mean > (g1_Q1 - 1.5*g1_IQR) & rats1$mean < (g1_Q3 + 1.5*g1_IQR))
rats2 <- subset(rats2, rats2$mean > (g2_Q1 - 1.5*g2_IQR) & rats2$mean < (g2_Q3 + 1.5*g2_IQR))
rats3 <- subset(rats3, rats3$mean > (g3_Q1 - 1.5*g3_IQR) & rats3$mean < (g3_Q3 + 1.5*g3_IQR))

glimpse(rats1)
glimpse(rats2)
glimpse(rats3)

rats_summary_out <- data.frame()
rats_summary_out <- rbind(rats_summary_out, rats1)
rats_summary_out <- rbind(rats_summary_out, rats2)
rats_summary_out <- rbind(rats_summary_out, rats3)
glimpse(rats_summary_out)

ggplot(rats_summary_out, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), Time (days) 1-64")

# Removing only the one outlier with the filter function.
rats_summary_out <- rats_summary %>% filter(mean <550)

ggplot(rats_summary_out, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), Time (days) 1-64")
```
  
As we previously saw from the data, group 1 has the lowest weight from the three groups and group 3 the highest. The weight difference between group 1 and 2-3 is quite large, so if lower weight was the goal, group 1 outperformed the rest with a large margin.

## Anova test

```{r}
# Creating the baseline from original data.
rats_base <- rats_long[rats_long$Time == 1,]

# Add the baseline from the original data as a new variable to the summary data
rats_summary_out <- rats_summary %>%
  mutate(baseline = rats_base$weight)
rats_summary_out

# Fit the linear model with the mean as the response
fit <- lm(mean ~ baseline, data = rats_summary_out)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```
  
According to the low P-value (1.57e-14), we can say that there is a difference in weight gain/loss between the three groups. So we made the correct first predictions with  the original plots. 

# Part 2  

## Linear mixed effects model  

```{r}
dim(bprs_long)
str(bprs_long)
glimpse(bprs_long)

ggplot(bprs_long, aes(x = week, y = bprs, col = subject)) +
  geom_line() +
  facet_grid(cols = vars(treatment), 
             labeller = label_both) +
  labs(title = "Effects of treatments 1 and 2 on bprs") +
  theme(legend.position = "top")
```
  At a first glance, it is quite difficult to say which treatment has better results.  

## Multiple linear regression model

```{r}
bprs_reg <- lm(bprs ~ week + treatment, data = bprs_long)
summary(bprs_reg)
```
  
## Random intercept model

```{r}
bprs_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = bprs_long, REML = FALSE)
summary(bprs_ref)
```

## Random Intercept and Random Slope Model

```{r}
# Create a random intercept and random slope model.
bprs_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = bprs_long, REML = FALSE)

# Print a summary of the model.
summary(bprs_ref1)

# Perform an ANOVA test on the two models.
anova(bprs_ref1, bprs_ref)
```

## Random Intercept and Random Slope Model with interaction

```{r}
# create a random intercept and random slope model with the interaction
bprs_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = bprs_long, REML = FALSE)

# print a summary of the model
summary(bprs_ref2)

# perform an ANOVA test on the two models
anova(bprs_ref2, bprs_ref1)

# draw the plot of RATSL with the observed Weight values
ggplot(bprs_long, aes(x = week, y = bprs, colour = subject)) +
  facet_grid(cols = vars(treatment), 
             labeller = label_both) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "bprs score") +
  theme(legend.position = "top")

# Create a vector of the fitted values
bprs_fitted <- fitted(bprs_ref2)


# Create a new column fitted to RATSL
bprs_long <- mutate(bprs_long, fitted = bprs_fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(bprs_long, aes(x = week, y = bprs_fitted, colour = subject)) +
  facet_grid(cols = vars(treatment), 
           labeller = label_both) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "Fitted bprs score") +
  theme(legend.position = "top")
```
  
It is really difficult to say which treatment had better results. If I had to make a quess, I would say that treatment 1 had a slightly better performance since most of the subjects seem to have a lower score in the end than the ones in treatment group 2.