# Chapter 5 - Dimensionality reduction techniques  
## Exploring the dataset

```{r}
# Loading packages.
library(dplyr)
library(MASS)
library(tidyr)
library(corrplot)
library(ggplot2)
library(GGally)

# Loading the dataset.
human <- read.csv("data/human.csv")

# Exploring structure and dimensions.
str(human)
dim(human)

# Graphical overview of the data.
pairs(human, pch = 20)

summary(human)

# Plot correlation between variables.
p <- ggpairs(human, upper = list(continuous = wrap("cor",size=2)))
p

# Create the correlation matrix and plot it.
cor_matrix <- round(cor(human),digits=5)
cor_matrix
corrplot(cor_matrix, method="circle")
```
  The dataset consists of 155 rows and 8 columns, being mostly numerical type (except GNI is integer type).  
  Female to male education ratio has a high positive correlation with Edu.Exp and Life.Exp and a high negative correlation with GNI, Mat.Mor and Ado.Birth. Highly educated women are less likely to die from childbirth and have a higher overall life expectancy.  
  Suprisingly, female to male labour force participation does not correlate heavily with any of the variables, neither do the GNI or Parli.F.
  
## Principal component analysis (PCA)

```{r}
# Performing PCA with raw human data
pca_human <- prcomp(human)

# Drawing a biplot of the principal component representation and the original variables.
biplot(pca_human, choices = 1:2)

```

## PCA with standardized variables
```{r}
# Standardizing the data.
human_std <- scale(human)

# Performing the PCA
pca_human_std <- prcomp(human_std)

# Drawing the biplot
biplot(pca_human, choices = 1:2)
```

The resulting plots seem to be identical.  
(Unfortunately due toa work trip and presentation planning I didn't have time to complete all of the exercises this week.)